Governing through patronage: the bargain for education in decentralized Brazil

### 2020-10-01 paper/paper.tex
update angle?

Across the world, vulnerable populations rely on governments for access to basic public services such as education. In the developing world, the quality of public education remains low: functional illiteracy is widespread, basic mathematical operations beyond the reach of most. This ``learning poverty'' is in  How can local governments improve the quality the of public services they deliver? In particular, how do political actors reshape bureaucratic institutions, and what are their downstream consequences for student learning.

To answer these questions, over two years I conducted fieldwork research in municipalities across Brazil. In interviews with local politicians, school principals and teachers, I discovered what many scholars working on the developing world already know: clientelism is pervasive, and it extends to public education with nefarious consequences \citep{stokes_brokers_2013}. However, the type of clientelism I observed was not embedded in electoral politics \citep{oliveros_making_2016} or partisan networks \citep{akhtari_political_2015,colonnelli_patronage_2017}, but the elite politics of government \citep{raile_executive_2011}. In Brazil, the exchange of public sector appointments for political favors -- a canonical example of patronage -- grants the mayor a carte blanche to enact her policy agenda.

>>>>

Across the world, vulnerable populations rely on governments for access to basic public services such as education. In the developing world, and in particular Latin America, the quality of public education remains low: functional illiteracy is widespread, basic mathematical operations beyond the reach of most \citep{vegas_raising_2007}. The World Bank refers to as ``learning poverty'': low gains from education even when access to a public school exists, in contrast with earlier focus on access to education.\footnote{For more information, see \hyperlink{here}{https://www.worldbank.org/en/topic/education/brief/learning-poverty}.} These outcomes are the end-result of a series of interlocking steps: from personnel decisions to service provision \citep{pepinsky_bureaucracy_2017}, each link in this institutional chain of command connects political principals to the child sitting in the classroom.

How can we intervene in this chain to improve the quality of public services? Previous studies have explored how political incentives can improve the quality of public goods provision. Some have focused on the direct link between political principal and bureaucratic agent: increasing political oversight over bureaucracies may improve public services \citep{gulzar_politicians_2017, raffler_does_2016}. Others have focused on the more indirect electoral accountability link binding politicians to voters, exploring how to strengthen it to induce changes  \citep{boas_electoral_2019, besley_political_2003} Whether explicitly or implicitly, these theories posit a unitary political actor who interacts with either the bureaucracy or the electorate. Yet in much of the developing world, these politicians do not govern alone: in fact, they are embedded in democratic institutions where multiplicity of elected officials and power-sharing arrangements dramatically reshape the institutional chain connecting politician to public service provision \citep{francois_how_2015}.

To shed light on this complexity, over two years I conducted fieldwork research in municipalities across Brazil. In interviews with local politicians, school principals and teachers, I discovered what many scholars working on the developing world already know: clientelism is pervasive, and it extends to public education with nefarious consequences \citep{stokes_brokers_2013}. However, the type of clientelism I observed was not embedded in electoral politics \citep{oliveros_making_2016} or partisan networks \citep{akhtari_political_2015,colonnelli_patronage_2017}, but the elite politics of government \citep{raile_executive_2011}. In Brazil, the exchange of public sector appointments for political favors -- a canonical example of patronage -- grants the mayor a carte blanche to enact her policy agenda.
